Motor Trends : Automatic or Manual transmission for better mileage ?
========================================================

**by P. Paquay**

## Executive summary

In this report we try to answer the question : "Is automatic or manual transmission better for mpg ?". To answer this question we used a dataset from the 1974 Motor Trend US magazine, and ran some statistical tests and a regression analysis. On one hand the statistical tests show (without controlling for other car design features) a difference in mean of about 7 miles more for the manual transmitted cars. On the other hand, the regression analysis indicate that, given that weight and 1/4 mile time are held constant, manual transmitted cars are 14.079 - 4.141 * weight miles per gallon better than automatic transmitted cars on average and also that this result is significant. So, we may conclude that lighter cars are better off with a manual transmission, but heavier cars are better off with an automatic one.

## Cleaning data

The first step of our analysis is simply to load and take a look at the data.

```{r}
data(mtcars)
str(mtcars)
```

Now we coerce the "cyl", "vs", "gear", "carb" and "am" variables into factor variables.

```{r}
mtcars$cyl <- factor(mtcars$cyl)
mtcars$vs <- factor(mtcars$vs)
mtcars$gear <- factor(mtcars$gear)
mtcars$carb <- factor(mtcars$carb)
mtcars$am <- factor(mtcars$am)
```

For a better readability, we rename the levels of the "am" variable into "Auto" and "Manual".

```{r}
levels(mtcars$am) <- c("Auto", "Manual")
```

## Exploratory analysis

We begin by plotting boxplots of the variable "mpg" when "am" is "Auto" or "Manual" (see Figure 1 in the appendix). This plot hints at an increase in mpg when gearing was manual but this data may have other variables which may play a bigger role in determination of "mpg".

We then plot the relationships between all the variables of the dataset (see Figure 2 in the appendix). We may note that variables like "wt", "cyl", "disp" and "hp" seem highly correlated together.

## Inference

We may also run some tests to compare the mpg means between automatic and manual transmissions.

### T-test

We begin by using a t-test assuming that the mileage data has a normal distribution.

```{r, results = 'hide'}
t.test(mpg ~ am, data = mtcars)
```

The p-value of `r t.test(mpg ~ am, data = mtcars)$p.value` clearly shows that the manual and automatic transmissions are significatively different.

### Wilcoxon test

Next we use a nonparametric test to determine if there's a difference in the population means.

```{r, results = 'hide'}
wilcox.test(mpg ~ am, data = mtcars)
```

Here again the p-value of `r wilcox.test(mpg ~ am, data = mtcars)$p.value` allow us to reject the null hypothesis that the mileage data of the manual and automatic transmissions are from the same population (indicating a difference).

## Regression analysis

First we need to select a model, we proceed by using the Bayesian Information Criteria (BIC) in a stepwise algorithm. This algorithm does not evaluate the BIC for all possible models but uses a search method that compares models sequentially. Thus it bears some comparison to the classical stepwise method but with the advantage that no dubious p-values are used.

```{r results = 'hide'}
model.all <- lm(mpg ~ ., data = mtcars)
n <- nrow(mtcars)
model.init <- step(model.all, direction = "backward", k = log(n))
```

```{r}
summary(model.init)$coefficients
```

The BIC algorithm tells us to consider "wt" and "qsec" as confounding variables. The individual p-values allows us to reject the hypothesis that the coefficients are null. The adjusted r-squared is `r summary(model.init)$adj.r.squared`, so we may conclude that more than `r round(summary(model.init)$adj.r.squared * 100)`% of the variation is explained by the model.

However, if we take a look a the scatter plot of "mpg" vs. "wt" by transmission type (see Figure 3 in the appendix) we may notice that the "wt" variable depends on whether or not the car is automatic transmitted (as automatic transmitted cars tend to weigh more than manual transmitted ones). Apparently, manual transmission only confers an advantage to lighter cars. If the car is heavier than approximately 3 tons, an automatic transmission is actually more fuel-efficient than a manual one. This fact suggests that it would be appropriate to include an interaction term between "wt" and "am".

```{r}
model <- lm(mpg ~ wt + qsec + am + wt:am, data = mtcars)
summary(model)$coefficients
```

The adjusted r-squared is now `r summary(model)$adj.r.squared`, so we may conclude that more than `r round(summary(model)$adj.r.squared * 100)`% of the variation is explained by the model. We will choose this model as our final model.

```{r}
anova <- anova(lm(mpg ~ am, data = mtcars), lm(mpg ~ am + wt, data = mtcars), model.init, model)
cbind(anova[1], anova[2], anova[3], anova[4], anova[5], anova[6])
```

We may notice that when we compare the model with only "am" as independant variable and our chosen model, we reject the null hypothesis that the variables "wt", "qsec" and "wt:am" don't contribute to the accuracy of the model.

The regression suggests that, "wt" and "qsec" variables remaining constant, manual transmitted cars can drive `r summary(model)$coef[4]` + `r summary(model)$coef[5]` * "wt" more miles per gallon on average than automatic transmitted cars, and the results are statistically significant. This means that for example, a 1000lbs manual transmitted car can drive `r summary(model)$coef[4] + summary(model)$coef[5] * 1` more miles per gallon than a same weight automatic transmitted one with the same 1/4 mile time and a a 4000lbs manual transmitted car can drive `r -(summary(model)$coef[4] + summary(model)$coef[5] * 4)` less miles per gallon than a same weight automatic transmitted one with the same 1/4 mile time.

Examining these coefficients allows us to determine exactly the point at which the fuel efficieny plots for automatic versus manual cars intersect, this point occurs at a weight of `r (-summary(model)$coef[4] / summary(model)$coef[5]) * 1000`lbs. We can explain this fact by noticing that almost all of the manual transmission cars are quite small. Therefore, even though cars with automatic transmission might get better mileage across almost all weights, the sample of manual cars consists almost solely of those weights in which manual cars win out.

```{r}
confint(model)[c(4, 5), ]
```

More accurately, we are 95% confident that the difference in miles per gallon between manual and automatic transmitted cars lies somewhere in the interval [`r as.numeric(confint(model)[4, ][1])` + `r as.numeric(confint(model)[5, ][1])` * wt, `r as.numeric(confint(model)[4, ][2])` + `r as.numeric(confint(model)[5, ][2])` * wt].

## Residuals and diagnostics

### Residual analysis

We begin by studying the residual plots (see Figure 4 in the appendix). These plots allow us to verify some assumptions made before.

1. The Residuals vs Fitted plot seem to verify the independance assumption as the points are randomly scattered on the plot (a Durbin-Watson test further confirms this assumption at the 0.05 level).
2. The Normal Q-Q plot seem to indicate that the residuals are normally distributed as the points hug the line pretty closely (a Shapiro-Wilk test further confirms this assumption at the 0.05 level).
3. The Scale-Location plot seem to verify the constant variance assumption as the points fall in a constant band (a Breusch-Pagan test further confirms this assumption at the 0.05 level).

### Leverages

We begin by computing the leverages for the "mtcars" dataset.

```{r}
leverage <- hatvalues(model)
```

Are any of the observations in the dataset outliers ? We find the outliers by selecting the observations with a hatvalue > 0.5.

```{r}
leverage[which(leverage > 0.5)]
```

### Dfbetas

Next we look at the Dfbetas of the observations.

```{r}
influential <- dfbetas(model)
```

Are any of the observations in the dataset influential ? We find the influential observations by selecting the ones with a dfbeta > 1 in magnitude.

```{r}
influential[which(abs(influential) > 1)]
```

## Appendix

### Figure 1 : Boxplots of "mpg" vs. "am"

```{r  fig.height = 10, fig.width = 10}
plot(mpg ~ am, data = mtcars)
title(main = "Mpg by transmission type", xlab = "am", ylab = "mpg")
```

### Figure 2 : Pairs graph

```{r  fig.height = 10, fig.width = 10}
pairs(mtcars, panel = panel.smooth, main = "Pairs graph for MTCars")
```

### Figure 3 : Scatter plot of "mpg" vs. "wt" by transmission type

```{r  fig.height = 10, fig.width = 10}
plot(mtcars$wt, mtcars$mpg, col = mtcars$am, pch = 19, xlab = "weight", ylab = "mpg")
title(main = "Scatter plot of mpg vs. wt by am")
legend("topright", c("automatic", "manual"), col = 1:2, pch = 19)
```

### Figure 4 : Residual plots

```{r fig.height = 10, fig.width = 10}
par(mfrow = c(2, 2))
plot(model)
```